---
title: "Problem Set 2 - Carter Hardy"
output: html_document
date: "2025-02-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1
#### Load the dataset
```{r}

birth <- read.table("/Users/carterhardy/spring 2025 UU/Prob-and-Stats-BMI6106/Problem set 2/birthwt.txt", header = TRUE, sep = " ")  

```

### A. How many observations are in the dataset?
#### I can see when I load in the data that there are 189 observations

### B. Examine each column

### C. How many individuals older than 30 smoke?
```{r}
#Filter the age of observations to those above 30
filtered_birth = birth[birth$age > 30, ]

#out of those above the age 30, sum of the ones who smoke
sum(filtered_birth$smoke == 1)
```
### D. plot a histogram of birth weights
```{r}
#this code creates a histogram of birth weights
hist(birth$bwt,
     main = "Histogram of Birth Weight",
     xlab = "Birth Weight (bwt) in Grams")
```

### G. Calculate the probability of randomly selecting an individual that has either a low birth weight or a mother who was a smoker. 

#### In order to find this, we will calculate P(A or B) = P(A) + P(B) - P(A & B)

```{r}

#total number of observations
n <- nrow(birth)

#prob of low birth weight
p_low = sum(birth$low == 1) / n

#prob of a mother who was a smoker
p_smoker = sum(birth$smoke == 1) / n

#prob of low birth weight and mother who was a smoker
p_both = sum(birth$smoke == 1 & birth$low == 1) / n

#P(A or B) = P(A) + P(B) - P(A & B)
p_union = p_low + p_smoker - p_both

p_union


```

### H. Calculate the probability of randomly selecting an individual that is white and has more than 3 physician visits during the first trimester.
```{r}
#P(A & B) 

#probability of both (individual is white and has more than 3 visits during first tri) divided by the total number of observations 
joint <- sum(birth$race == 1 & birth$ftv > 3) / n

joint
```








