---
title: "Clustering Homework - Due April 9th"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Carter Hardy

The Breast Cancer dataset, also known as the Wisconsin Diagnostic Breast Cancer (WDBC) dataset, contains measurements from digitized images of fine needle aspirate (FNA) biopsies of breast masses. Each sample is described by 30 numeric features representing the characteristics of the cell nuclei, such as radius, texture, perimeter, area, and smoothness. The dataset includes 569 samples, each labeled as either malignant (M) or benign (B).

Your job today is use the learned clustering methods in class to best describe this dataset. The Ultimate goal is to understand the classificatory power of this dataset, which variables are the most important at separating the two classes and to statistically compare multiple clustering methods.

## 1. Load Data 

```{r}
data <- read.csv("Breast_Cancer.csv")
head(data)

# Load libraries
library(ggplot2)
library(factoextra)
library(tidyverse)
library(ggplot2)
library(caret)
```

## 2. PCA (10)

- Scree plot
- Biplot
- Variance explained - which variables contribute the most on PC1,2?
- Confusion matrix
```{r}
#run PCA
pca <- prcomp(data[,2:31], scale. = TRUE)

#Scree plot
fviz_eig(pca, addlabels = TRUE, ylim = c(0, 100)) +
  ggtitle("Scree Plot - Variance Explained by Each PC")
#The elbow is around the 3 PC

#Biplot
fviz_pca_biplot(pca,
                label = "var",   # show variable vectors
                habillage = data$diagnosis, # color by diagnosis
                addEllipses = TRUE,
                col.var = "black") +
  ggtitle("PCA Biplot with diagnosis")

#Variance explained (How much each variable contributes to each PC)
loadings <- as.data.frame(pca$rotation)
print(loadings)
##Variable Contributions to Each PC (%)
# Contributions (percent)
contrib <- get_pca_var(pca)$contrib
round(contrib, 2)

#Confusion matrix
# We'll use the first 3 principal components for clustering
pca_data <- as.data.frame(pca$x[, 1:3])
#Run K-means on the PCA-transformed data
set.seed(49)
kmeans_result <- kmeans(pca_data, centers = 2,nstart = 25)

pca_data$Cluster <- as.factor(kmeans_result$cluster)
pca_data$diagnosis <- data$diagnosis

###plot PCA with cluster assignments
ggplot(pca_data, aes(PC1, PC2, color = Cluster, shape = diagnosis)) +
  geom_point(size = 3) +
  labs(title = "PCA + K-means Clustering on diagnosis Data")


##Confusion matrix (compare clusters with actual species)
table(Cluster = kmeans_result$cluster, diagnosis = data$diagnosis)

cluster_labels <- ifelse(kmeans_result$cluster == 1, "M", "B")

# Now create factors with the same levels
predicted <- factor(cluster_labels, levels = c("B", "M"))
actual <- factor(data$diagnosis, levels = c("B", "M"))

con_pca = confusionMatrix(predicted, actual)
con_pca
#accuracy .9104
```
Our PCA confusion matrix shows that we have an accuracy of 91%.

## 3. t-SNE (10)

- Try different perplexities
- Visualize colored by true labels
- confusion matrix
```{r}
library(Rtsne)
#remove duplicate rows
data_unique <- data[!duplicated(data[,2:31]), ]

#perplexity 30
# Run t-SNE
tsne_result <- Rtsne(data_unique[,2:31], perplexity = 30)
tsne_data <- as.data.frame(tsne_result$Y)

set.seed(49)  # for reproducibility
kmeans_result <- kmeans(tsne_data, centers = 2)

# Actual labels
true_labels <- data_unique$diagnosis

# Predicted clusters
cluster_labels <- ifelse(kmeans_result$cluster == 1, "M", "B")

# Now create factors with the same levels
predicted_tsne <- factor(cluster_labels, levels = c("B", "M"))
actual_tsne <- factor(true_labels, levels = c("B", "M"))


# Create a confusion matrix
table(Cluster = cluster_labels, diagnosis = true_labels)

con_tsne = confusionMatrix(predicted_tsne, actual_tsne) 
con_tsne
#accuracy .819
```


With the perplexity of 30, our t-SNE confusion matrix shows we have an accuracy of 82%


```{r}
#perplexity 15
# Run t-SNE
tsne_result2 <- Rtsne(data_unique[,2:31], perplexity = 15)
tsne_data2 <- as.data.frame(tsne_result2$Y)

set.seed(49)  # for reproducibility
kmeans_result2 <- kmeans(tsne_data2, centers = 2)

# Actual labels
true_labels <- data_unique$diagnosis

# Predicted clusters
cluster_labels2 <- ifelse(kmeans_result2$cluster == 1, "M", "B")

# Now create factors with the same levels
predicted_tsne2 <- factor(cluster_labels2, levels = c("B", "M"))
actual_tsne2 <- factor(true_labels, levels = c("B", "M"))


# Create a confusion matrix
table(Cluster = cluster_labels2, diagnosis = true_labels)

con_tsne2 = confusionMatrix(predicted_tsne2, actual_tsne2) 
con_tsne2
#accuracy .61
```
With the perplexity of 15, our t-SNE confusion matrix shows we have an accuracy of 62%, this is worse than our accuracy with a perplexity of 30.


## 4. K-means Clustering (10)

- Elbow method
- Silhouette score
- Compare to true labels - confusion matrix
```{r}
library(cluster)
library(factoextra)
library(purrr)
library(caret)
#Elbow method
###How many Ks?
data_wo <- scale(data[, -1])

#Elbow method
wss <- function(k) {
  kmeans(data_wo, k, nstart = 10)$tot.withinss
}

k.values <- 1:15
wss_values <- map_dbl(k.values, wss)

plot(k.values, wss_values,
     type = "b", pch = 19, frame = FALSE,
     xlab = "Number of clusters K",
     ylab = "Total within-cluster sum of squares",
     main = "Elbow Method for Optimal K")

#K-means with k = 2
set.seed(123)
kmeans_result_elbow <- kmeans(data_wo, centers = 2, nstart = 25)
clusters_elbow <- kmeans_result_elbow$cluster

#Silhouette score
dist_matrix <- dist(data_wo, method = "euclidean")
sil <- silhouette(clusters_elbow, dist_matrix)
fviz_silhouette(sil)

# Step 5: Confusion matrix (comparing clusters to actual diagnosis)
# Convert clusters to "M"/"B" labels — you may need to flip them
cluster_labels_elbow <- ifelse(clusters_elbow == 1, "M", "B")

# Match factors
predicted_elbow <- factor(cluster_labels_elbow, levels = c("B", "M"))
actual_elbow <- factor(data$diagnosis, levels = c("B", "M"))

con_kmean = confusionMatrix(predicted_elbow, actual_elbow)
con_kmean
#accuracy .9104
```
Our Kmeans confusion matrix shows that we have an accuracy of 91%.

## 5. Hierarchical Clustering (10)

- Try different linkage methods
- Plot dendrograms
- Evaluate - confusion matrix
```{r}
#I had to test it on a small sample because at first I could not get it to load.

#set.seed(49)
#sample_idx <- sample(1:nrow(data_wo), 500)
#data_small <- data_wo[sample_idx, ]
#data_scale <- scale(data_small)
#dist_data <- dist(data_scale, method = "euclidean")
#library(factoextra)
#hc <- hclust(dist_data, method = "ward.D2")
#fviz_dend(hc, k = 3, rect = TRUE, main = "Ward.D2 Dendrogram", cex = 0.6)


data_scale = scale(data_wo)
dist_data <- dist(data_scale, method = "euclidean")

##different linkage methods
library(gridExtra)
#Try several linkage methods
methods <- c("average", "single", "complete", "ward.D2")

#Plot dendrograms
plots <- list()
for (m in methods) {
  hc <- hclust(dist_data, method = m)
  plots[[m]] <- fviz_dend(hc, k = 3, rect = TRUE, main = paste("Dendrogram -", m), cex = 0.6)
}

# Show 2x2 grid
grid.arrange(grobs = plots, ncol = 2)

##Use one of the clustering methods to cut the tree
hc_ward <- hclust(dist_data, method = "ward.D2")
cluster_labels_hier <- cutree(hc_ward, k = 2)

table(cluster_labels_hier, data$diagnosis) 


predicted_hier <- factor(ifelse(cluster_labels_hier == 1, "M", "B"), levels = c("B", "M"))
actual_hier <- factor(data$diagnosis, levels = c("B", "M"))

#confusion matrix
con_hier = confusionMatrix(predicted_hier, actual_hier)
con_hier
#accuracy .8805
```
Our Hierarchical Clustering confusion matrix shows that we have an accuracy of 88%.


## 6. Combination of Methods (10)

- Does combining methods work better?
Based on the accuracy of the confusion matrix below, by combining t-SNE and hierarchical clustering, this did not work better than Kmeans or PCA. Sometime combining methods may work better, but in this case the combination of PCA and Kmeans worked better than t-SNE and hierarchical clustering.

- Evaluate - confusion matrix

```{r}
#Combining methods, T-sne and hierarchical 
library(Rtsne)
#remove duplicate rows
data_unique <- data[!duplicated(data[,2:31]), ]
diagnosis_unique <- data_unique$diagnosis

#Run t-SNE
tsne_result <- Rtsne(data_unique[,2:31], perplexity = 30)
tsne_data <- as.data.frame(tsne_result$Y)

#start hierarchical  clustering
dist_data_combined <- dist(tsne_data, method = "euclidean")
hc_ward_comb <- hclust(dist_data_combined, method = "ward.D2")
cluster_labels_comb <- cutree(hc_ward_comb, k = 2)

#check table
table(cluster_labels_comb, diagnosis_unique) 

#create confusion matrix
predicted_comb <- factor(ifelse(cluster_labels_comb == 1, "M", "B"), levels = c("B", "M"))
actual_comb <- factor(diagnosis_unique, levels = c("B", "M"))

#confusion matrix
con_combine = confusionMatrix(predicted_comb, actual_comb)
con_combine
#accuracy .819
```
By combing methods we achieved an accuracy of 82%, shown by our confusion matrix.

## 7. Evaluation (20)

- Confusion matrices comparison and analysis (why are these different?)
```{r}
#Compare confusion matrices 
con_pca
#accuracy .9104

con_tsne #with 30 perplexity 
#accuracy .819

con_tsne2 #with 15 perplexity
#accuracy .61

con_kmean
#accuracy .9104

con_hier
#accuracy .8805

con_combine
#accuracy .819
```
Each confusion matrix has a different method that is giving its predictions and then comparing the actual results from the data. From these confusion matrices we can compare and analyze the difference in the accuracy, the successful predictions, failed predictions, and more from the different methods.
We can see that PCA and Kmeans gave us our highest accuracy, and combining t-SNE and hierarchical clustering, and t-SNE both with 30 and 15 perplexities gave us our lowest accuracy. t-SNE with 15 perplexity gave us the lowest accuracy with 62%.


```{r}
library(clusterCrit)
library(cluster)

# Pre-allocate dataframe for storing metrics
metrics_df <- data.frame(
  Method = character(),
  Dunn = numeric(),
  DB = numeric(),
  Avg_Silhouette = numeric(),
  stringsAsFactors = FALSE
)

# Loop through clustering methods
for (m in methods) {
  # Perform hierarchical clustering using the chosen method
  hc <- hclust(dist_data, method = m)
  
  # Cut the tree to create clusters
  labels <- cutree(hc, k = 3)
  
  # Calculate silhouette score
  sil <- silhouette(labels, dist_data)
  
  # Calculate Dunn and Davies-Bouldin indices
  crit_vals <- intCriteria(as.matrix(data_scale), labels, c("Dunn", "Davies_Bouldin"))
  
  # Append metrics for this method to the dataframe
  metrics_df <- rbind(metrics_df, data.frame(
    Method = m,
    Dunn = round(crit_vals$dunn, 3),
    DB = round(crit_vals$davies_bouldin, 3),
    Avg_Silhouette = round(mean(sil[, 3]), 3)
  ))
}

# Print the metrics for all methods
print(metrics_df)

# High Dunn, low DB, and high silhouette = best clustering method.

```
- Dunn index (why are these different?) 
**Answer:** higher Dunn index tells us that the data is well seperated and better clustering, these scores for the 4 different methods are relatively low scores, with single method being the best but only being .5.
- Davies-Bouldin index (why are these different?)
**Answer:** a lower index is better for the Davies-Bouldin index, sinfle has the lowest index score with .38 while the others are higher.
From these scores we can take away that these clusters are not great.

## 8. Conclusion (20)
- Overall conclusion of your analysis one paragraph 200 words +/- 10 words

Using PCA, t-SNE, Kmeans, hierarchical clustering, and then combining t_SNE and hierarchical, we could find varying levels of accuracy of the predicted outcome of breast cancer, benign or malignant, against the actual outcome of breast cancer. Some of the methods worked slightly better than the others, but our lowest accuracy was 62% from t-SNE with 15 perplexities, which is still considered okay. In contrast, our higher accuracy methods, like PCA and Kmeans, have an accuracy of 91%. The other techniques fall within 82%-89%. Overall, these methods were all able to predict the correct type of breast cancer over half of the time, with most showing good accuracy. This analysis was an effective way of clustering different types of breast cancer. I found PCA to be my preferred method for this analysis. With my statistical background, I am familiar with the elbow plot and choosing k or the number of PCs based on the elbow graph. Deciding where the cut-off is can be intimidating, but with continued practice, selecting the elbow cut-off becomes another skill. I found hierarchical clustering the most intimidating and challenging in this assignment because my computer struggled to process the dendrograms; once I swept my environment, it could handle it.

## 9. Additional Questions (10 - 5 each)

- How many principal components are required to explain 80% of the variance?
**Answer:** For our PCA, if we add the first 3 PCs, it would explain 79% of the variance. So to answer the question, 4 PCs would get us above 80%, but 3 PCs is really close to 80% of the variance.

- True or False: t-SNE preserves global distances between samples.
**Answer:** False

- Why do we scale the variables in this dataset?
**Answer:** We scale the variables in this dataset in order to calculate distance, like the euclydian distance, for methods that rely on distance calculations, which calculations like kmeans and silhouette rely on. Scaling ensures that each feature is contributes equally to distance based calculations.

- Which metric favors high separation and low intra-cluster spread?.

  A.  Dunn Index
  B.  DB index
  C.  Silhouette
  D.  Euclidean Distance
**Answer:** A. Dunn Index
```